## Parsing FastQ files

For a more comprehensive review of what can be performed using `R/Bioconductor` 
for short reads let us have a look at some data from the public domain. The 
Short Read Archive (SRA) hosted by the NBCI and provides a public repository 
where high dimensional data can be shared with the community. Study `ERX337002` 
looks at the metagenomics of food (http://www.ncbi.nlm.nih.gov/sra/ERX337002).
The Ion Torrent platform has been used to sequence the microbial content of an artisanal cheese. 

The raw data has been downloaded from the public domain and has been converted 
from the `sra` format into a `fastq` file format (using SRA Toolkit) that can be 
used with a number of packages. The fastq file prepared has been gzipped to 
save space (and increase read performance).

```
curl -0 ftp://ftp-trace.ncbi.nlm.nih.gov/sra/sra-instant/reads/ByExpr/sra/ERX/ERX337/ERX337002/ERR364233.sra
fastq-dump ERR364233.sra
gzip ERR364233.fastq
```

A brief summary of the content in the file can be prepared by loading the fastq 
file with the `readFastq` function.

```{r}
cheese <- readFastq(file.path(NGS_DIR, "ERR364233.subset.fastq.gz"))
print(cheese)
```

```{r}
qaSummary <- qa(cheese, NGS_DIR)
report(qaSummary, dest=".")
```

```{block, type="rmdexercise"}
1. Enter the above commands.
2. Click on the `Files` tab in the bottom-right corner of RStudio.
3. Navigate to your working directory (this should be the RNA-seq_analysis project folder).
4. Click on `index.html` and select `View in Web Browser` (if a popup window appears click Try again).
5. Have a look a the report from the `qaSummary()` function.
```


This shows that there are 200K sequence reads in the sequence collection 
that have a length of between 8 and 358 nucleotides. Let's have a look at the distribution of read lengths:

```{r}
hist(width(cheese), breaks=100)
```

We recommend that we do not use such short reads in our sequence collection.
Filter out sequences that are less than 100 nt in length and replot the distribution.

```{r}
cheese.filtered <- cheese[-which(width(cheese) < 100)]
hist(width(cheese.filtered), breaks=100)
```

```{block, type="rmdexercise"}
How many reads were removed from the filtering step? ______________________
(**hint** `length()`)
```

We have looked at the distribution of sequence lengths in the collection. We can
also check the distribution of individual nucleotides in the collection. 
The `alphabetFrequency` function collects frequency counts for each of the 
IUPAC nucleotides in a `DNAStringSet` object. The method will generate a table 
that be summarised and plotted using your preferred graphics library.

```{r}
freq <- alphabetFrequency(sread(cheese.filtered))
totalCount <- colSums(freq)
barplot(totalCount[1:4])
```

```{block, type="rmdexercise"}
1. Run the command above.
2. What *data type* is the variable `freq`? ________________
3. What are the dimensions of `freq`? ______________________
4. Why did we only plot the first 4 elements of `totalCount`? _________________


```

Similar plots can be prepared for di- and trinucleotide frequencies, this looks 
at all possible words. You can similarly look for hexanucleotide frequencies 
or any other number, check the `alphabetFrequency()` documentation. This can
assists in detecting any over-represented subsequences.

```{r}
barplot(colSums(dinucleotideFrequency(sread(cheese.filtered))))
```


Depending on the sequencing strategy that you applied, you may have ligated 
synthetic adapter sequences, primers and other constructs within the target 
amplicons. There are a number of methods for stripping such synthetic sequences 
from your sequence collection.

```{r}
head(sread(cheese.filtered))
cloningPrimer <- "AACATTTGTCCTAATCATTGGGGATA"
cheese.clipped <- trimLRPatterns(Lpattern = cloningPrimer, subject=cheese.filtered,
                                 max.Lmismatch = 0.33)
head(sread(cheese.clipped))
```

There is of course a lot more that can be done to allow for perfect, imperfect and 
truncated primer sequences or adapters at either the forward or reverse ends of the sequence.

One of the reasons why the `fastq` file format has become so popular in genomics is 
that in addition to storing information relating to the DNA sequence, the file stores
information on the DNA sequence and the quality (confidence score that the base has been measured correctly). 

The quality scores in a ShortRead data collection are accessible using the `quality()` function.

```{r}
head(quality(cheese.clipped))
```

```{r}
encoding(quality(cheese.clipped))
```

```{r}
cheese.quality <- as(quality(cheese.clipped), "matrix")
hist(rowMeans(cheese.quality[sample(nrow(cheese.quality), 25000),],na.rm=TRUE), main="", xlab="")
```

```{block, type="rmdexercise"}
1. Type in the commands above.
2. What *data type* is the variable `cheese.quality`? ____________________________
3. What is the dimension of `cheese.quality`? _________________________________
4. How does the number of columns relate to the length of the reads? _____________
  
5. Below are the broken-down parts of the `hist(...)` function that produced
   the plot above. The list starts with the most inner function and works it's
   way out. Can you describe what each inner function is doing and the expected result?
   The letter in the curly braces e.g. \{a\} means substitute the result
   from the part \{a\} into this function.
   
   a. `nrow(cheese.quality)`: _________________________________________________
   b. `sample({a}, 25000)`  : _________________________________________________
   c. `cheese.quality[{b},]` : ________________________________________________
   d. `rowMeans({c}, na.rm=TRUE)`: ____________________________________________
   e. `hist({d}, main="", xlab="")`: ___________________________________________
```

The code above presents a text format (representation of the integer based quality scores) 
of the quality data. This is converted into a qualityMatrix that can then be
plotted to show the distribution of quality scores across the sequence collection.

Plotting the per-base quality scores is the goal of many software applications 
and this can be managed simply in R as well. One of the requirements to produce 
these plots is however, to reduce the dimensions of the data. Plotting the 
characteristics for every read is unnecessary since a sample should convey 
trends within the data.

```{r}
boxplot(cheese.quality[sample(nrow(cheese.quality), 25000), 1:100],out.cex=0.5)
```


```{r}
mung <- function(i) {
  lower <- i
  upper <- i + 4
  if (upper > ncol(cheese.quality)) {
    upper <- ncol(cheese.quality)
  }
  return(rowMeans(cheese.quality[,seq(lower,upper)], na.rm=TRUE))
}

sequence <- seq(1, ncol(cheese.quality), 5)
groupedQuality <- as.data.frame(sapply(sequence, mung))
colnames(groupedQuality) <- as.character(sequence)

boxplot(groupedQuality[sample(nrow(groupedQuality), 25000),], outcex=0.5)
abline(h=c(20,28),lty=2,col='white')
rect(-5,28,358,45,col=rgb(0.1,1,0,alpha=0.25),border=NA)
rect(-5,20,358,28,col=rgb(1,0.5,0,alpha=0.25),border=NA)
rect(-5,0,358,20,col=rgb(1,0.6,0.6,alpha=0.25),border=NA)
```

Having prepared this box-and-whisker plot we can see that the distribution of 
quality scores is generally OK but there are quite a large number of low quality 
bases described as the interquartile ranges within the whiskers. To make the 
data most valuable to the analysis it would be worthwhile to filter the sequences 
to exclude the substrings of lowest quality from the 3'-end forwards. The 
`ShortRead` package implements some easy to use functions such as `trimTails` and 
`trimTailw`.

```{r}
cheese.clipped <- trimTailw(cheese.clipped, k=4, a="4", halfwidth=5)
cheese.quality <- as(quality(cheese.clipped), "matrix")
dim(cheese.quality)
sequence <- seq(1, ncol(cheese.quality), 5)
groupedQuality <- as.data.frame(sapply(sequence, mung))
colnames(groupedQuality) <- as.character(sequence)
boxplot(groupedQuality[sample(nrow(groupedQuality), 25000),], outcex=0.5)
abline(h=c(20,28),lty=2,col='white')
rect(-5,28,358,45,col=rgb(0.1,1,0,alpha=0.25),border=NA)
rect(-5,20,358,28,col=rgb(1,0.8,0,alpha=0.25),border=NA)
rect(-5,0,358,20,col=rgb(1,0.6,0.6,alpha=0.25),border=NA)
```

This shows that we have greatly improved the overall structure of the data and 
have removed a large number of the less-than-perfect bases. This example may 
have been a little more aggressive than we would really wish to apply in the laboratory setting.

The `ShortRead` package has a well implemented framework for filtering 
sequences and it would be simple to implement other filters as required. The 
filter can be illustrated by filtering out the homopolymers from the sequence 
collection. These provide mechanisms for restricting the sequence collection 
on the basis of e.g. base composition, number of N-residues and even relative abundance.

A whole lot more can be done using the `ShortRead` and `Biostrings` packages. 

```{block, type="rmdexercise"}
**Prepare some QC information for an Illumina data collection**

Using the RNA-Seq reads that you will perform differential expression with,
prepare a brief report of the data. Using the `qaSummary` function to prepare
a synopsis of the library content and investigate the per-base quality
scores across the sequence collection. Would you recommend that we re-run
the analysis after some data trimming?

Have a look at the over-represented k-mer words in the sequence collection 
and see if you can create a filter to strip-out the sequence that contain the
most abundant k-mer.
```

